---
title: "Class 08: Breast Cancer Mini Project"
author: "Brian Wong (PID: A18639001)"
format: pdf
toc: TRUE
---

## Background

In today's class we will apply the methods and techniques clustering and PCA to help make sense of a real world breast cancer FNA biopsy data set. 

## Data Import

We start by importing our data. It is a CSV file so we will use the `read.csv()` function.

```{r}
# Save your input data file into your Project directory
fna.data <- "WisconsinCancer.csv"

# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(fna.data, row.names=1)

head(wisc.df, 4)
```

Make sure to remove the first `diagnosis` column - I don't want to use this for my machine learning models. We will use it later to compare our results to the expert diagnosis.

```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]

# Create diagnosis vector for later 
diagnosis <- wisc.df$diagnosis 
```

## Exploratory Data Analysis

> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.df)
```

> Q2. How many of the observations have a malignant diagnosis?

```{r}
sum(diagnosis == "M")
```

> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
length(grep("_mean", colnames(wisc.data)))
```

## Principal Component Analysis

The main function here is `prcomp()` and we want to make sure we set the optional argument `scale = TRUE`:

### Performing PCA
```{r}
# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)
```
```{r}
# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data, scale = TRUE)
```

```{r}
# Look at summary of results
summary(wisc.pr)
```
> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

44.27% of the original variance is captured by the first principal component (PC1)

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

The first 3 PCs add up to 72.64% of the original variance. Hence 3 PCs are required to describe at least 70% of the original variance in the data.

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

The first 7 PCs add up to 91.01% of the original varaiance. Hence 7 PCs are required to describe at least 90% of the original variance in the data.

### Interpreting PCA results

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

The plot is extremely difficult to understand. The labels are overlapping with each other and is unreadable. Also the data overlaps with each other and is not visible for us to interpret.

Our main PCA "score plot" or "PC plot" of results"
```{r}
# Scatter plot observations by components 1 and 2
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

The plots slightly differ in how the points are spread. On the PC2 versus PC3 scale, it seems that the points on the PC2 scale are spread out more while the points on the PC3 scale are slightly more packed.

```{r}
# Repeat for components 1 and 3
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()

```

### Variance Explained

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

```{r}
## ggplot based graph
#install.packages("factoextra")
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```

### Communicating PCA Results

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

The component of the loading vector for the feature concave.points_mean is -0.2608538. There are no other features with larger contriutions than this one.
```{r}
ggplot(wisc.pr$rotation) + aes(PC1, (rownames(wisc.pr$rotation))) + geom_col()

wisc.pr$rotation["concave.points_mean",1]
wisc.pr$rotation[,1]
```

## Hiearchical Clustering

First scale the data (with the `scale`() function), then calculate a distance matrix (with the `dist()` function). Then cluster with `hclust()` function and plot.

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)
```

```{r}
data.dist <- dist(data.scaled)
```

```{r}
wisc.hclust <- hclust(data.dist, method = "complete")
```

### Results of Hiearchical Clustering

> Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

Just right below height 20, is when the clustering model has 4 clusters.

```{r}
plot(wisc.hclust)
abline(h=19, col="red", lty=2)
```

### Selecting Number of Clusters

You can also use `cutree()` function with a argument `k=4` rather than `h = height`

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)

table(wisc.hclust.clusters, diagnosis)
```
> Q11. OPTIONAL: Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 6? How do you judge the quality of your result in each case?

There are different definitions of quality that may vary between clustering.

### Using Different Methods

> Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

Using "ward.D2" gives the favored result in my opinion. The first few clusters are easy to see and there is a clear difference in Height among them as opposed to methods like "single".

```{r}
wisc.hclust2 <- hclust(data.dist, method = "ward.D2")
plot(wisc.hclust2)
```


## Combining Methods

Here we will take our PCA results and use those as input for clustering. In other words our `wisc.pr$x` scores that we plotted above (the main output from PCA - how the data lie on our new principal component axis/variables) and use a subset of the PCs that capture the most variance as input for `hclust()`

### Clustering PCA Results

```{r}
pc.dist <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(pc.dist, method = "ward.D2")
plot(wisc.pr.hclust)
```

Cut the dendrogram/tree into two main groups/clusters:

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```
I want to know how the clustering in `grps` with values of 1 or 2 correspond to the expert `diagnosis`

```{r}
table(grps, diagnosis)
```

My clustering **group 1** are mostly "M" diagnosis (179) and my clustering **group 2** are mostly "B" diagnosis.

24 False Positives (FP)
179 True Positives (TP)
333 True Negatives (TN)
33 False Negatives (FN)

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2) +
  geom_point(col=grps)
```

> Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?

There are 24 False Positives (FP), 179 True Positives (TP), 333 True Negatives (TN), 33 False Negatives (FN) in this newly created hclust model with 2 clusters of "M" and "B" diagnoses.

```{r}
table(grps, diagnosis)
```
> Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.

There are more groups in the original cluster with few individuals in the 2 separate groups. There seems to be relatively less False Positives in group 1, but slightly more false negatives in group 2.

```{r}
table(wisc.hclust.clusters, diagnosis)
table(grps, diagnosis)
```
## Sensitivity/Specificity

Sensitivity TP/(TP+FN)
```{r}
179/(179+33)
```

Specificity TN/(TN+FP)
```{r}
333/(333 + 24)
```


## Prediction 

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```
```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```

> Q16. Which of these new patients should we prioritize for follow up based on your results?

We should prioritize these new patients in group 2 for a follow up.
