---
title: "Class 18: Pertussis Mini Project"
author: "Brian Wong (PID: A18639001)"
format: pdf
toc: TRUE
---

## Background
Pertussis (whooping cough) is a common lung infection caused by the bacteria B. Pertussis. It can infect anyone but is most deadly for infants (under 1 year of age).

# CDC Tracking Data

The CDC tracks the number of of Pertussis cases:

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

:::{.callout-note collapse = "true"}

```{r}
cdc <- data.frame(
  year = c(1922L,1923L,1924L,1925L,
           1926L,1927L,1928L,1929L,1930L,1931L,
           1932L,1933L,1934L,1935L,1936L,
           1937L,1938L,1939L,1940L,1941L,1942L,
           1943L,1944L,1945L,1946L,1947L,
           1948L,1949L,1950L,1951L,1952L,
           1953L,1954L,1955L,1956L,1957L,1958L,
           1959L,1960L,1961L,1962L,1963L,
           1964L,1965L,1966L,1967L,1968L,1969L,
           1970L,1971L,1972L,1973L,1974L,
           1975L,1976L,1977L,1978L,1979L,1980L,
           1981L,1982L,1983L,1984L,1985L,
           1986L,1987L,1988L,1989L,1990L,
           1991L,1992L,1993L,1994L,1995L,1996L,
           1997L,1998L,1999L,2000L,2001L,
           2002L,2003L,2004L,2005L,2006L,2007L,
           2008L,2009L,2010L,2011L,2012L,
           2013L,2014L,2015L,2016L,2017L,2018L,
           2019L,2020L,2021L,2022L,2023L,2024L, 2025L),
  cases = c(107473,164191,165418,152003,
            202210,181411,161799,197371,
            166914,172559,215343,179135,265269,
            180518,147237,214652,227319,103188,
            183866,222202,191383,191890,109873,
            133792,109860,156517,74715,69479,
            120718,68687,45030,37129,60886,
            62786,31732,28295,32148,40005,
            14809,11468,17749,17135,13005,6799,
            7717,9718,4810,3285,4249,3036,
            3287,1759,2402,1738,1010,2177,2063,
            1623,1730,1248,1895,2463,2276,
            3589,4195,2823,3450,4157,4570,
            2719,4083,6586,4617,5137,7796,6564,
            7405,7298,7867,7580,9771,11647,
            25827,25616,15632,10454,13278,
            16858,27550,18719,48277,28639,32971,
            20762,17972,18975,15609,18617,
            6124,2116,3044,7063,22538,21996)
)
```
:::

```{r}
library(ggplot2)

ggplot(cdc) + 
  aes(year, cases) + 
  geom_point() + 
  geom_line() + 
  labs(x = "Year", y = "Number of cases", title = "Pertussis Cases by Year (1922 - 2025)")
```

> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

> Q. Add annotation lines for the major milestones of wP vaccination roll-out (1946) and the switch to the aP vaccine (1996).

```{r}
ggplot(cdc) + 
  aes(year, cases) + 
  geom_point() + 
  geom_line() + 
  labs(x = "Year", y = "Number of cases", 
       title = "Pertussis Cases by Year (1922 - 2025)") +
  geom_vline(xintercept = 1946, col = "blue", lty = 2) + 
  geom_vline(xintercept = 1996, col = "red", lty = 2) + 
  geom_vline(xintercept = 2020, col = "gray", lty = 2)
```

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine, the number of cases of Pertussis decreased a significant amount. Making the aP vaccine safer and more effective with less side effects than the wP vaccine could have caused less cases and people to be more confident in the vaccine. After a while, it seems like the number of cases started to rise again, which could be due to how the weakening of the aP vaccine with less side effects and an increased number of testing. 


## Exploring CMI-PB Data

The CMI-PB project's < https://www.cmi-pb.org/ > is to provide the scientific community with a comprehensive, high-quality and freely accessible resource of Pertussis booster vaccination.

Basically, they make available a large dataset on the immune response to Pertussis. They use a "booster" vaccination as a proxy for Pertussis infection.

They make their data available as a JSON format API. We can read this into R with the `read_json()` function from the **jsonlite** package:

```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", 
                     simplifyVector = TRUE)
head(subject)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

> How many aP and wP individual are there in this `subject` table?

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

> Q. Is this representative of the US population?

No, this is not representative of the US population. 


Working with Dates:

```{r}
library(lubridate)
```

```{r}
today()

today() - ymd("2000-01-01")
```

```{r}
time_length( today() - ymd("2000-01-01"),  "years")
```

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}
library(dplyr)

subject$age <- today() - ymd(subject$year_of_birth)

ap <- subject |> filter(infancy_vac == "aP")

round( summary( time_length( ap$age, "years" ) ) )
```

```{r}
wp <- subject %>% filter(infancy_vac == "wP")
round( summary( time_length( wp$age, "years" ) ) )
```

```{r}
t.test(ap$age, wp$age)
```

> Q8. Determine the age of all individuals at time of boost?

```{r}
int <- ymd(subject$date_of_boost) - ymd(subject$year_of_birth)
age_at_boost <- time_length(int, "year")
head(age_at_boost)
```

> Q9. With the help of a faceted boxplot or histogram (see below), do you think these two groups are significantly different?

Yes, these two groups are significantly different.

```{r}
ggplot(subject) +
  aes(time_length(age, "year"),
      fill=as.factor(infancy_vac)) +
  geom_histogram(show.legend=FALSE) +
  facet_wrap(vars(infancy_vac), nrow=2) +
  xlab("Age in years")
```

```{r}
x <- t.test(time_length( wp$age, "years" ),
       time_length( ap$age, "years" ))

x$p.value
```

We can read more tables from the CMI-PB database

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5_1/specimen",
                      simplifyVector = TRUE)
ab_titer <- read_json("https://www.cmi-pb.org/api/v5_1/plasma_ab_titer",
                      simplifyVector = TRUE)
```

```{r}
head(specimen)
```

```{r}
head(ab_titer)
```

To make sense of all this data, we need to "join" (a.k.a. merge or "link") all of these tables together. Only then will you know that a given Ab measurement (from the `ab_titer` table) was collected on a certain date (from the `specimen` table) from a certain wP or aP subject (from the `subject` table).

We can use **dplyr** and the `*.join()` family of functions to do this.

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r}
library(dplyr)
meta <- inner_join(subject, specimen)
dim(meta)
head(meta)
```

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

Let's do one more `inner_join()` to join the `ab_titer` with all our `meta` data.

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```

> Q11. How many different Ab "isotypes" values are in this dataset?

```{r}
table(abdata$isotype)
```

> Q12. How many different "antigen" values are measured?

```{r}
table(abdata$antigen)
```

Let's focus on IgG isotype

```{r}
igg <- abdata |> filter(isotype == "IgG")

head(igg)
```

Let's make a plot of `MFI_normalised` values for all `antigen` values.

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) + 
  geom_boxplot()
```

The antigens "PT", "FIM2/3", and "FHA" appear to have the widest range of values.

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens:

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() + 
    xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```

> Q14. Is there a difference for these responses between aP and wP individuals?

The antigens "PT", "FIM2/3", and "FHA" appear to have the widest range of values.

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) + 
  geom_boxplot()
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  facet_wrap(vars(visit), nrow=2) +
  xlim(0,75) +
  theme_bw()
```

```{r}
igg %>% filter(visit != 8) %>%
ggplot() +
  aes(MFI_normalised, antigen, col=infancy_vac ) +
  geom_boxplot(show.legend = FALSE) + 
  xlim(0,75) +
  facet_wrap(vars(infancy_vac, visit), nrow=2)
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) + 
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```

Is there a difference with time (i.e. before booster shot vs after booster shot)?

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) + 
  geom_boxplot() +
  facet_wrap(~visit)
```

> Q15. Filter to pull out only two specific antigens for analysis and create a boxplot for each. You can chose any you like. Below I picked a “control” antigen (“OVA”, that is not in our vaccines) and a clear antigen of interest (“PT”, Pertussis Toxin, one of the key virulence factors produced by the bacterium B. pertussis).

```{r}
filter(igg, antigen=="OVA") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
filter(igg, antigen=="PT") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

```{r}
filter(igg, antigen=="FIM2/3") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()
```

> Q16. What do you notice about these two antigens time courses and the PT data in particular?

The two antigens have their levels increase over the course of the visits. In the PT data, PT levels rise and exceed OVA antigen levels. The trend follows similarly between wP and aP subjects.

> Q17. Do you see any clear difference in aP vs. wP responses?

There does not seem to be a clear difference in aP vs. wP response as the ranges and medians seem to be close to each other.


```{r}
# Filter to 2021 Dataset

ab.PT.21 <- abdata |> filter(dataset == "2021_dataset", 
                              isotype == "IgG",  antigen == "PT")
  
ggplot(ab.PT.21) +
  aes(x=planned_day_relative_to_boost,
      y=MFI_normalised,
      col=infancy_vac,
      group=subject_id) +
  geom_point() +
  geom_line(alpha = 0.2) +

  # Line of Fit for aP and wP groups
  geom_smooth(aes(group = infancy_vac, col = infancy_vac), 
              se = FALSE, 
              span = 0.3) + 
  
  geom_vline(xintercept=0, linetype="dashed") +
  geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)") 
```

> Q18. Does this trend look similar for the 2020 dataset?

In general, yes, the general trend looks similar to the 2020 dataset. However, there are a few differences with the 2020 dataset that are quite apparent such as the peaks between the "aP" and "wP" in the first 14 days. 

```{r}
# Filter to 2020 Dataset

ab.PT.20 <- abdata |> filter(dataset == "2020_dataset", 
                              isotype == "IgG",  antigen == "PT")
  
ggplot(ab.PT.20) +
  aes(x=planned_day_relative_to_boost,
      y=MFI_normalised,
      col=infancy_vac,
      group=subject_id) +
  geom_point() +
  geom_line(alpha = 0.2) +
  geom_vline(xintercept=0, linetype="dashed") +
  geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2020 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)") 
```

## Obtaining CMI-PB RNASeq Data

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
```

```{r}
#meta <- inner_join(specimen, subject)
ssrna <- inner_join(rna, meta)
```

> Q19. Make a plot of the time course of gene expression for IGHG1 gene (i.e. a plot of visit vs. tpm).

```{r}
ggplot(ssrna) +
  aes(visit, tpm, group=subject_id) +
  geom_point() +
  geom_line(alpha=0.2)
```

> Q20.: What do you notice about the expression of this gene (i.e. when is it at it’s maximum level)?

The expression of the gene was at its maximal level on the 4th day of visit and slowly reduced over the course of the next two days. One subject seemed to linearly decrease expression while a majority of subjects decreased expression levels exponentially.

> Q21. Does this pattern in time match the trend of antibody titer data? If not, why not? 

Yes, this pattern in time matches the trend of antibody titer data. From the boxplot data above, we can see that the titer levels increased over a few days until week 4 and then progressively lowered. 


```{r}
ggplot(ssrna) +
  aes(tpm, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(vars(visit))
```

```{r}
ssrna %>%  
  filter(visit==4) %>% 
  ggplot() +
    aes(tpm, col=infancy_vac) + geom_density() + 
    geom_rug() 
```


